Working report

### 2018-06-01 DraftReport.tex
code and files added for time series of ADA,KRA,HO

	\section{Extreme analysis}
	\subsection{Ground work}
	In order to use the extreme value theory, in our case \textbf{POT} method, we nee our data to be stationary, to ensure identical, and independent. To make data behave independent, we remove the smaller of the two esceedances, if they are closer then a certain number of days from each other. This lag is usually set to 3 or 5 days, depending on region and time of year. To make sure that the exceedances follows a \textbf{Poisson point process},the standard way is to consider data in cluster of a specific size and then only take the maximum if more then one observation within that cluster exceeds the threshold. This makes it important to pick a high enough threshold, since else a lot of our observations will be lost.
	To ensure stationarity is more diffucult and several approaches are used. One is to divide the data into seasons and use different thresholds for each season. The issue with this method is that we assume stationarity within each season and therefore can get unrealistic return values if there in fact is a trend in the data. If one knows that the data depends on other factors, such as the yield depends on temperature and rain, one can use a Box-Cox transformation on the data and then let the drift and trend parameter depend on these other paramters, so called \textbf{covariates}. This however requires that we know of these covariates, can estimate or already have models for how they affect our data and have measurments of these covariates over the same time period as our own data.\\

>>>>

	\section{Extreme analysis}
	\subsection{Ground work}
	In order to use the extreme value theory, in our case \textbf{POT} method, we need our data to be stationary, to ensure identical, and independent. To make data behave independent, we remove the smaller exceedances, if they are more then one closer then a certain number of days from each other. This lag is usually set to 3 or 5 days, depending on region and time of year. To make sure that the exceedances follows a \textbf{Poisson point process},the standard way is to consider data in cluster of a specific size and then only take the maximum if more then one observation within that cluster exceeds the threshold. This makes it important to pick a high enough threshold, since else a lot of our observations will be lost.
	To ensure stationarity is more difficult and several approaches are used. One is to divide the data into seasons and use different thresholds for each season. The issue with this method is that we assume stationarity within each season and therefore can get unrealistic return values if there in fact is a trend in the data. If one knows that the data depends on other factors, such as the yield depends on temperature and rain, one can use a Box-Cox transformation on the data and then let the drift and trend parameter depend on these other paramters, so called \textbf{covariates}. This however requires that we know of these covariates, can estimate or already have models for how they affect our data and have measurments of these covariates over the same time period as our own data.\\
	To investigate stationarity in our time serie, we first applied a decomposition on both the full time series and then on each season. This to look at trends both on an annual time scale but also on a seasonal one. By decomposing even the seasons we can remove influence on seasonality within each season, i.e we usually have a lot more rain in September then in November. To statisically check for stationarity, we used the \textbf{Augmented dickey-Fuller test} (ADF). Many uses \textbf{ManKendell}, but since this test only uses non ties and we have loads of zeroes in several seasons, it will not be very reliable (?).

	\subsubsection{Time series analysis}
	{\Large\textbf{HO}}: Figure \ref{tsHO} shows the trends for years and seasons. The only season that is showing some sort of trend is DJF, which seems to have a slight increasing trend. However, the scale is so small,that it is more or less neglectibe. SON is not showing a clear trend, but an increased volatility in the trend after 1995. There is a big volatilty in the JJA trend, suggesting that the rain in that season depends on some other process. MAM is the most steady season. All time series rejects the null hypothesis for the ADF test, hence suggesting that there is no unit root, i.e stationary. 
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.6\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsHO.pdf}
			\caption{Full time series}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=\linewidth]{tsHOmam.pdf}
			\caption{MAM}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsHOjja.pdf}
			\caption{JJA}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=\linewidth]{tsHOson.pdf}
			\caption{SON}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsHOdjf.pdf}
			\caption{DJF}
		\end{subfigure}%
		\caption{Decomposition of time series for HO}
		\label{tsHO}
	\end{figure}

	{\Large\textbf{ADA}}: In figure \ref{tsADA} we can once again see that there is no clear trend on the annual time series, which the ADF test confirms. MAM has again the smoothest trend series with only one big dip around 1988, and there seems to be a decrease in the most extreme values. JJA is similarly to HO showing periodic behaviour indicating that there is some underlying process effecting the rain in this season. Cannot really detect a trend. SON shows a fairly flat trend graph, but an increase in extreme observations after some dry years between 1990-2000. DJF is also showing an increse in the amplitude of extreme values even if the number of extreme values is not clearly increasing. The tend graph is not flat mostly due to the fact that we nearly only have zero values.
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.6\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsADA.pdf}
			\caption{Full time series}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=\linewidth]{tsADAmam.pdf}
			\caption{MAM}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsADAjja.pdf}
			\caption{JJA}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=\linewidth]{tsADAson.pdf}
			\caption{SON}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsADAdjf.pdf}
			\caption{DJF}
		\end{subfigure}%
		\caption{Decomposition of time series for ADA}
		\label{tsADA}
	\end{figure}

	{\Large\textbf{KRA}}: Comparing fig \ref{tsKRA} with the two previous, we can see a lot of similarities. There is once again no trend in the full time series and it is somewhere between the behaviour of HO and ADA, with HO being much smoother and ADA more volitile. But for this station MAM and JJA are similarly smooth. MAM seems to have an increase in amplitude in extreme values, which is not visible in JJA. SON instead looks like it is having less exteme values, wheras DJF clearly has an increase both in number and amplitude in extreme values. All time series rejects the null hypothesis for the ADF test.
	\begin{figure}[H]
		\centering
		\begin{subfigure}{0.6\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsKRA.pdf}
			\caption{Full time series}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=\linewidth]{tsKRAmam.pdf}
			\caption{MAM}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsKRAjja.pdf}
			\caption{JJA}
		\end{subfigure}
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=\linewidth]{tsKRAson.pdf}
			\caption{SON}
		\end{subfigure}%
		\begin{subfigure}{0.5\textwidth}
			\centering
			\includegraphics[width=1\linewidth]{tsKRAdjf.pdf}
			\caption{DJF}
		\end{subfigure}%
		\caption{Decomposition of time series for KRA}
		\label{tsKRA}
	\end{figure}

### 2018-06-01 WorkingReport.tex
code and files added for time series of ADA,KRA,HO

	\section{West African metorology}
	The monsoon differs over Africa due to its geographical difference. In west Africa, the northern part consists of land and the southern part of ocean wheras east Africa only consists of land, eventhough the northern part is wider then the southern. In the northern hemisphere summer a thermal low pressure builds up over the continent around 20\textdegree N and the ITCZ moves northward to 15\textdegree N. At the same time, heat troughs develop over North Atlantic ocean, the anticyclone system \textbf{St Helena} builds up ove the South Atlantic ocean and cold water is flowing northward along the south-west African coast. Beacause of all this a pressure gradient is formed between South Atlantic and north Africa which makes the south-easterly flows to recurve and become the south-westerly monsoon flow over west Africa.(McGregor:1998). This flow brings cold and moist oceanic air which is the reason for rain, so the area with maximum rainfall is where we have thickest air masses, i.e closest to the coast. North of the ITCSZ conditions are genrally cloudless and dry and south of the zone we find the most cloud cover and maximum rain. Dry years are coused by: retardation of the northward movement of the south-west monsoon, a southward displacemnt of the intertropical discontinuity,the near-equatorial trough and the xone of maximum surface pressure and anoalously cold water to the nort-west of the line linking SW West Africa nd NE Brasil. 
	During the northern hemisphere winter, West Africa is instead under the influence of  north-easterly trade winds which bring dry and stable air masses, often containing dust particles. These winds are called \textbf{Harmattan}. At this time, the ITCZ is located to the south of the southern west-African coast, hence why there is no rain over west Africa this time of year.

	It is not clear wheather ENSO has an impact on the African monsoon or not, but it seems that very intense warm phases of it can reduce the monsoon precipitation, as recorded during the -83 ENSO. The \textbf{west African mid-Troposhperic jet} is a thermal wind that occurce due to the temperature gradient between the warm Sahara and the cold Gulf of Guinea. We can find maximum rainfall on the equatorward side of the jet. The \textbf{ITCZ} is composed of two zones, one with low pressure, maximum surface temperature and wind confluence and another zone with maximum cloudiness, therefore also maximum rainfall and wind converge. These two zones can be as far aprat as 1000 km. 
	\textbf{Squall lines} is a linear system of many thunderstorms or clouds that behave as one and therefore can live for 13-15 h instead of just a couple of hourse. They can be hundreds of km long and 30 km wide and are characterised by their explosive growth, rapid propagation and their convex leading ege(?). They peak in the afternoon and usually form west of mountains which implies that surface heating and orographic  effects creates them. In west Africa, the African easterly jet seems to have a big influence on the initiation of them. Moisture in the atmosphere is a key thing for creating squall lines, by having moist lower levels and dry mid levels a postive feedback is created over the zone and therfore further developes the zone. There are more squall lines during the beginning and end of the monsoon because the dry mid level is absent during the middle of the monsoon.

@@

	\section{Dictionary}
	\begin{itemize}
		\item \textbf{Convergence}: Difference in wind speed forces air to "pile up", which creates a vertical movemnet upwards or downwards depending if the convergence is on the surface or up in the atmosphere.
		\item \textbf{Confluence}: Difference in wind speed gets air together but as wind enters the confluence zone, it speeds up, hence does not converge.
		\item \textbf{Baroclinic}: A atmosphere for which the density depends on both temperature and pressure.
		\item \textbf{Barotropic}: A atmosphere for which the density only depends on pressure.
		\item \textbf{Wind shear}: Change in wind speed and/or direction over a short distance in the atmosphere. It can be both horisontaly or verticaly and usually observed close to weather fronts or thermal winds.
		\item \textbf{Trough (dal)}: Extended region of relatively low atmospheric pressure.
		\item \textbf{Latent heat}: Energy transfer to the atmosphere due to evaporation.
		\item \textbf{Relative humidity}: ratio of amount of water present it the atmosphere realtive to the amount that could be present at that given temperature.
		\item \textbf{Dew formation(dagg)}: Condensation on cool surfaces.
		\item \textbf{Anabatic}: Upward wind motion.
		\item \textbf{Adiabatic}: Any change in internal energy only depends on work such as compression or expansion. Opposite is \textbf{diabatic} which is a change in energy between a system and its surrounding due to a temperature gradient.
		\item \textbf{Sensitivity}: An evaluation of how much each input contributes to the model output uncertainty. This can be done in numerous of ways, linear regression being one or running the model and changing one variable at a time.
	\end{itemize}

>>>>

	\section{West African metorology}
	The monsoon differs over Africa due to its geographical difference. In west Africa, the northern part consists of land and the southern part of ocean wheras east Africa only consists of land, eventhough the northern part is wider then the southern. In the northern hemisphere summer a thermal low pressure builds up over the continent around 20\textdegree N and the ITCZ moves northward to 15\textdegree N. At the same time, heat troughs develop over North Atlantic ocean, the anticyclone system \textbf{St Helena} builds up ove the South Atlantic ocean and cold water is flowing northward along the south-west African coast. Beacause of all this a pressure gradient is formed between South Atlantic and north Africa which makes the south-easterly flows to recurve and become the south-westerly monsoon flow over west Africa.(McGregor:1998). This flow brings cold and moist oceanic air which is the reason for rain, so the area with maximum rainfall is where we have thickest air masses, i.e closest to the coast. North of the ITCSZ conditions are generally cloudless and dry and south of the zone we find the most cloud cover and maximum rain. Dry years are caused by: retardation of the northward movement of the south-west monsoon, a southward displacement of the intertropical discontinuity,the near-equatorial trough and the zone of maximum surface pressure and anomalously cold water to the nort-west of the line linking SW West Africa and NE Brasil. 
	During the northern hemisphere winter, West Africa is instead under the influence of  north-easterly trade winds which bring dry and stable air masses, often containing dust particles. These winds are called \textbf{Harmattan}. At this time, the ITCZ is located to the south of the southern west-African coast, hence why there is no rain over west Africa this time of year.

	It is not clear wheather ENSO has an impact on the African monsoon or not, but it seems that very intense warm phases of it can reduce the monsoon precipitation, as recorded during the -83 ENSO. The \textbf{west African mid-Troposhperic jet} is a thermal wind that occurce due to the temperature gradient between the warm Sahara and the cold Gulf of Guinea. We can find maximum rainfall on the equatorward side of the jet. The \textbf{ITCZ} is composed of two zones, one with low pressure, maximum surface temperature and wind confluence and another zone with maximum cloudiness, therefore also maximum rainfall and wind converge. These two zones can be as far aprat as 1000 km. 
	 In west Africa, the African easterly jet seems to have a big influence on the initiation of them. Moisture in the atmosphere is a key thing for creating squall lines, by having moist lower levels and dry mid levels a postive feedback is created over the zone and therfore further developes the zone. There are more squall lines during the beginning and end of the monsoon because the dry mid level is absent during the middle of the monsoon.

@@

	\section{Dictionary}
	\begin{itemize}
		\item \textbf{Convergence}: Difference in wind speed forces air to "pile up", which creates a vertical movemnet upwards or downwards depending if the convergence is on the surface or up in the atmosphere.
		\item \textbf{Confluence}: Difference in wind speed gets air together but as wind enters the confluence zone, it speeds up, hence does not converge.
		\item \textbf{Baroclinic}: A atmosphere for which the density depends on both temperature and pressure.
		\item \textbf{Barotropic}: A atmosphere for which the density only depends on pressure.
		\item \textbf{Wind shear}: Change in wind speed and/or direction over a short distance in the atmosphere. It can be both horisontaly or verticaly and usually observed close to weather fronts or thermal winds.
		\item \textbf{Trough (dal)}: Extended region of relatively low atmospheric pressure.
		\item \textbf{Latent heat}: Energy transfer to the atmosphere due to evaporation.
		\item \textbf{Relative humidity}: ratio of amount of water present it the atmosphere realtive to the amount that could be present at that given temperature.
		\item \textbf{Dew formation(dagg)}: Condensation on cool surfaces.
		\item \textbf{Anabatic}: Upward wind motion.
		\item \textbf{Adiabatic}: Any change in internal energy only depends on work such as compression or expansion. Opposite is \textbf{diabatic} which is a change in energy between a system and its surrounding due to a temperature gradient.
		\item \textbf{Sensitivity}: An evaluation of how much each input contributes to the model output uncertainty. This can be done in numerous of ways, linear regression being one or running the model and changing one variable at a time.
		\item \textbf{Squall lines} is a linear system of many thunderstorms or clouds that behave as one and therefore can live for 13-15 h instead of just a couple of hourse. They can be hundreds of km long and 30 km wide and are characterised by their explosive growth, rapid propagation and their convex leading ege(?). They peak in the afternoon and usually form west of mountains which implies that surface heating and orographic  effects creates them
	\end{itemize}
